Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-16 talk/dls2012/paper.tex

\subsection{Python}
The Python interpreter of the RPython framework is a complete Python
version 2.7 compatible interpreter. A set of numerical
calculations were implemented in both Python and in C and their
runtimes are compared in Figure~\ref{fig:benchmarks}.\footnote{
    The benchmarks and the scripts to run them can be found in the repository for this paper:
    \texttt{https://bitbucket.org/pypy/extradoc/src/ tip/talk/dls2012/benchmarks}
}

The benchmarks are
\begin{itemize}
\item {\bf sqrt}: approximates the square root of $y$. The approximation is 
initiated to $x_0=y/2$ and the benchmark consists of a single loop updating this
approximation using $x_i = \left( x_{i-1} + y/x_{i-1} \right) / 2$ for $1\leq i < 10^8$. 
Only the latest calculated value $x_i$ is kept alive as a local variable within the loop.
There are three different versions of this benchmark where $x_i$
  is represented with different type of objects: int's, float's and
  Fix16's. The latter, Fix16, is a custom class that implements
  fixpoint arithmetic with 16 bits precision. In Python there is only
  a single implementation of the benchmark that gets specialized
  depending on the class of it's input argument, $y$, while in C,
  there are three different implementations.
\item {\bf conv3}: one-dimensional convolution with fixed kernel-size $3$. A single loop
is used to calculate a vector ${\bf b} = \left(b_1, \cdots, b_n\right)$ from a vector
${\bf a} = \left(a_1, \cdots, a_n\right)$ and a kernel ${\bf k} = \left(k_1, k_2, k_3\right)$ using 
$b_i = k_3 a_i + k_2 a_{i+1} + k_1 a_{i+2}$ for $1 \leq i \leq n$. Both the output vector, $\bf b$, 
and the input vectors, $\bf a$ and $\bf k$, are allocated prior to running the benchmark. It is executed 
with $n=10^5$ and $n=10^6$.
\item {\bf conv5}: one-dimensional convolution with fixed kernel-size $5$. Similar to conv3, but with 
${\bf k} = \left(k_1, k_2, k_3, k_4, k_5\right)$. The enumeration of the elements in $\bf k$ is still 
hardcoded into the implementation making the benchmark consist of a single loop too.
\item {\bf conv3x3}: two-dimensional convolution with kernel of fixed
  size $3 \times 3$ using a custom class to represent two-dimensional
  arrays. It is implemented as two nested loops that iterates over the elements of the 
$n\times n$ output matrix ${\bf B} = \left(b_{i,j}\right)$ and calculates each element from the input matrix
${\bf A} = \left(a_{i,j}\right)$ and a kernel ${\bf K} = \left(k_{i,j}\right)$ using $b_{i,j} = $
\begin{equation}
  \label{eq:convsum}
  \begin{array}{lclclc}
    k_{3,3} a_{i-1,j-1} &+& k_{3,2} a_{i-1,j} &+& k_{3,1} a_{i-1,j+1} & + \\
    k_{2,3} a_{i,j-1}   &+& k_{2,2} a_{i,j}   &+& k_{2,1} a_{i,j+1}   & + \\
    k_{1,3} a_{i+1,j-1} &+& k_{1,2} a_{i+1,j} &+& k_{1,1} a_{i+1,j+1}  \\
  \end{array}
\end{equation}
for $1 \leq i \leq n$ and $1 \leq j \leq n$.
The memory for storing the matrices are again allocated outside the benchmark and $n=1000$ was used.
\item {\bf dilate3x3}: two-dimensional dilation with kernel of fixed
  size $3 \times 3$. This is similar to convolution but instead of
  summing over the terms in Equation~\ref{eq:convsum}, the maximum over those terms is taken. That places a
  external call to a max function within the loop that prevents some
  of the optimizations.
\item {\bf sobel}: a low-level video processing algorithm used to
  locate edges in an image. It calculates the gradient magnitude
  using sobel derivatives. A Sobel x-derivative, $D_x$, of a $n \times n$ image, ${I}$, is formed
by convolving ${I}$ with
\begin{equation}
  {K} = \left(
  \begin{array}{ccc}
    -1 & 0 & 1 \\
    -2 & 0 & 2 \\
    -1 & 0 & 1 \\
  \end{array}
  \right) ,
\end{equation}
and a Sobel y-derivative, $D_y$, is formed convolving $I$ with $K^\top$. The gradient magnitude is 
then formed for each pixel independently by $\sqrt{D_x^2 + D_y^2}$. The two convolutions and the pixelwise
magnitude calculation are combined in the implementation of this benchmark and calculated in a single pass over
the input image. This single pass consists of two nested loops with a somewhat larger amount of calculations 
performed each iteration as compared to the other benchmarks.
\end{itemize}

The sobel and conv3x3 benchmarks are implemented
on top of a custom two-dimensional array class.
It is
a straightforward implementation providing 2 dimensional
indexing with out of bounds checks. For the C implementations it is
implemented as a C++ class. The other benchmarks are implemented in
plain C. All the benchmarks except sqrt operate on C double-precision floating
point numbers, both in the Python and the C code.

>>>>

\subsection{Python}
The Python interpreter of the RPython framework is a complete Python
version 2.7 compatible interpreter. A set of numerical
calculations were implemented in both Python, C and Lua and their
runtimes are compared in Figure~\ref{fig:benchmarks}.\footnote{
    The benchmarks and the scripts to run them can be found in the repository for this paper:
    \texttt{https://bitbucket.org/pypy/extradoc/src/ tip/talk/dls2012/benchmarks}
}

The benchmarks are
\begin{itemize}
\item {\bf sqrt}$\left(T\right)$: approximates the square root of $y$. The approximation is 
initiated to $x_0=y/2$ and the benchmark consists of a single loop updating this
approximation using $x_i = \left( x_{i-1} + y/x_{i-1} \right) / 2$ for $1\leq i < 10^8$. 
Only the latest calculated value $x_i$ is kept alive as a local variable within the loop.
There are three different versions of this benchmark where $x_i$
  is represented with different type of objects, $T$,: int's, float's and
  Fix16's. The latter, Fix16, is a custom class that implements
  fixpoint arithmetic with 16 bits precision. In Python there is only
  a single implementation of the benchmark that gets specialized
  depending on the class of it's input argument, $y$, while in C,
  there are three different implementations.
\item {\bf conv3}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $3$. A single loop
is used to calculate a vector ${\bf b} = \left(b_1, \cdots, b_n\right)$ from a vector
${\bf a} = \left(a_1, \cdots, a_n\right)$ and a kernel ${\bf k} = \left(k_1, k_2, k_3\right)$ using 
$b_i = k_3 a_i + k_2 a_{i+1} + k_1 a_{i+2}$ for $1 \leq i \leq n$. Both the output vector, $\bf b$, 
and the input vectors, $\bf a$ and $\bf k$, are allocated prior to running the benchmark. It is executed 
with $n=10^5$ and $n=10^6$.
\item {\bf conv5}$\left(n\right)$: one-dimensional convolution with fixed kernel-size $5$. Similar to conv3, but with 
${\bf k} = \left(k_1, k_2, k_3, k_4, k_5\right)$. The enumeration of the elements in $\bf k$ is still 
hardcoded into the implementation making the benchmark consist of a single loop too.
\item {\bf conv3x3}$\left(n,m\right)$: two-dimensional convolution with kernel of fixed
  size $3 \times 3$ using a custom class to represent two-dimensional
  arrays. It is implemented as two nested loops that iterates over the elements of the 
$m\times n$ output matrix ${\bf B} = \left(b_{i,j}\right)$ and calculates each element from the input matrix
${\bf A} = \left(a_{i,j}\right)$ and a kernel ${\bf K} = \left(k_{i,j}\right)$ using $b_{i,j} = $
\begin{equation}
  \label{eq:convsum}
  \begin{array}{lclclc}
    k_{3,3} a_{i-1,j-1} &+& k_{3,2} a_{i-1,j} &+& k_{3,1} a_{i-1,j+1} & + \\
    k_{2,3} a_{i,j-1}   &+& k_{2,2} a_{i,j}   &+& k_{2,1} a_{i,j+1}   & + \\
    k_{1,3} a_{i+1,j-1} &+& k_{1,2} a_{i+1,j} &+& k_{1,1} a_{i+1,j+1}  \\
  \end{array}
\end{equation}
for $1 \leq i \leq m$ and $1 \leq j \leq n$.
The memory for storing the matrices are again allocated outside the benchmark and $(n,m)=(1000,1000)$ 
as well as $(n,m)=(1000000,3)$ was used.
\item {\bf dilate3x3}$\left(n\right)$: two-dimensional dilation with kernel of fixed
  size $3 \times 3$. This is similar to convolution but instead of
  summing over the terms in Equation~\ref{eq:convsum}, the maximum over those terms is taken. That places a
  external call to a max function within the loop that prevents some
  of the optimizations.
\item {\bf sobel}$\left(n\right)$: a low-level video processing algorithm used to
  locate edges in an image. It calculates the gradient magnitude
  using sobel derivatives. A Sobel x-derivative, $D_x$, of a $n \times n$ image, ${I}$, is formed
by convolving ${I}$ with
\begin{equation}
  {K} = \left(
  \begin{array}{ccc}
    -1 & 0 & 1 \\
    -2 & 0 & 2 \\
    -1 & 0 & 1 \\
  \end{array}
  \right) ,
\end{equation}
and a Sobel y-derivative, $D_y$, is formed convolving $I$ with $K^\top$. The gradient magnitude is 
then formed for each pixel independently by $\sqrt{D_x^2 + D_y^2}$. The two convolutions and the pixelwise
magnitude calculation are combined in the implementation of this benchmark and calculated in a single pass over
the input image. This single pass consists of two nested loops with a somewhat larger amount of calculations 
performed each iteration as compared to the other benchmarks.
\end{itemize}

The sobel and conv3x3 benchmarks are implemented
on top of a custom two-dimensional array class.
It is
a straightforward implementation providing 2 dimensional
indexing with out of bounds checks and 
data stored in row-major order.
For the C implementations it is
implemented as a C++ class. The other benchmarks are implemented in
plain C. All the benchmarks except sqrt operate on C double-precision floating
point numbers, both in the Python and the C code.

In addition we also ported the 
SciMark\footnote{\texttt{http://math.nist.gov/scimark2/}} benchmakts to python, and compared 
their runtimes with the already existing Lua and C implementations. 
This port was performed after the release of the pypy used to run the benchmarks which means that 
these benchmarks have not influenced the pypy implementation.
SciMark consists of

\begin{itemize}
\item {\bf SOR}$\left(n, c\right)$: Jacobi successive over-relaxation on a $n\times n$ grid repreated $c$ times.
The same custom two-dimensional array class as described above is used to represent
the gird.
\item {\bf SparseMatMult}$\left(n, z, c\right)$: Matrix multiplication between a $n\times n$ sparse matrix,
stored in compressed-row format, and a full storage vector, stored in a normal array. The matrix has $z$ non-zero elements and the calculation is repeated $c$ times.
\item {\bf MonteCarlo}$\left(n\right)$: Monte Carlo integration by generating $n$ points uniformly distributed over the unit square and computing the ratio of those within the unit circle.
\item {\bf LU}$\left(n, c\right)$: LU factorization of an $n \times n$ matrix. The rows of the matrix is shuffled which makes the previously used two-dimensional array class unsuitable. Instead a list of arrays is used to represent the matrix. The calculation is repeated $c$ times.
\item {\bf FFT}$\left(n, c\right)$: Fast Fourier Transform of a vector with $n$ elements, represented as an array, repeated $c$ times.
\end{itemize}
